New lower bounds for Schur and weak Schur numbers

### 2021-10-14 article/schur.tex
modifs :O

\begin{proposition}
	Let \(n \in [\![2, +\infty[\![\). We define \(S^+(n)\) as the greatest integer \(w\) such that there is a S-template with 
	width \(w\) and \(n\) colors. 
	\(S^+(n)\) is well defined and verifies
	\[
	2S(n-1)+1 \leqslant S^+(n) \leqslant S(n).
	\]
\end{proposition}

@@

\begin{theorem}
\label{thm:Stemp}
	Let \((p,k), (q,n) \in (\mathbb{N}^*)^2\). If there is a \(S\)-template with width \(q\) and \(n+1\) colors 
	and a sum-free \(k\)-partition of \([\![1,p]\!]\) then there is a partition of \([\![1,pq+m_{n+1}
	-1]\!]\) into \(n+k\) sum-free subsets. \(m_{n+1}\) denotes the minimum element colored with the special color in the \(S\)-template.
\end{theorem}

@@

\begin{proposition}
\label{prop:rafStemp}
Let \((q, n) \in \left(\mathbb{N}^* \right)^2\) and let \(f\) be a coloring associated to a S-template with width \(q\) and \(n+1\) 
colors. Let \(b \in \mathbb{N}\) and assume there is a coloring \(g\) of 
\([\![1, b]\!]\) with \(n+1\) colors such that:

@@

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the coloring \(x \longmapsto g(x - q S(n))\), we have
\[ S(n+k) \geqslant q S(k) + b.\]
\end{proposition}
This proposition corresponds to the fact that the hypotheses made on the coloring of the last row can be weakened because 
the constraints in the definition of a S-template prevent a number from: 
\begin{itemize}
\item being the sum of two of numbers of its own color. 
\item creating a sum in its own color. 
\end{itemize}
The later is not always relevant on the last row given that it is not completely filled.
As a result, we can change the coloring on \([\![ q S(k) + 1,  q S(k) + b]\!]\) (the last row) to extend the previous partitions.
\newline

@@

\begin{theorem}
	Let \((p,k), (q,n) \in (\mathbb{N}^*)^2\). If there is a S-template with width \(q\) and \(n+1\) colors,
	and a S-template with width \(p\) and \(k\) colors, then there also is a S-template with width \(pq\) and \((n+k)\) 
	colors.
\end{theorem}

@@

A S-template with \(n\) colors is said to be symmetric if the sum-free \(n\)-partition derived 
from this template by applying the extension procedure with a sum-free partition of length 1 is symmetric. 
\end{definition}

@@

\par
Inequalities (\ref{S(n+1)}), (\ref{S(n+2)}) and (\ref{S(n+3)}) cannot be further improved (with this definition of S-template). 
Inequality (\ref{S(n+4)}) cannot be further improved by only searching for symmetric S-templates whose special color is the 
last in the order of apparition (and with a multiplicative factor less than or equal to 118). Inequality (\ref{S(n+5)}) can most 
likely be further improved but the improvement probably will not be substantial. Finally, a S-template corresponding to inequality 
(\ref{S(n+6)}) was found by extending into a S-template the Schur 6 partition used in \cite{rowley2021improved} (owing to the 
size of this template, its interest is limited and since it can easily be derived from above mentioned partition, it is not given in 
\ref{S-templates}). Although we could not find a better S-template with seven colors, inequality (\ref{S(n+6)}) can definitely be 
improved by a wide margin.

>>>>

\begin{proposition}
	Let \(n \in [\![2, +\infty[\![\). We define \(S^+(n)\) as the greatest integer \(w\) such that there is an S-template with 
	width \(w\) and \(n\) colors. 
	\(S^+(n)\) is well defined and verifies
	\[
	2S(n-1)+1 \leqslant S^+(n) \leqslant S(n).
	\]
\end{proposition}

@@

\begin{theorem}
\label{thm:Stemp}
	Let \((p,k), (q,n) \in (\mathbb{N}^*)^2\). If there are a \(S\)-template with width \(q\) and \(n+1\) colors 
	and a sum-free \(k\)-partition of \([\![1,p]\!]\) then there is a partition of \([\![1,pq+m_{n+1}
	-1]\!]\) into \(n+k\) sum-free subsets. \(m_{n+1}\) denotes the minimum element colored with the special color in the \(S\)-template.
\end{theorem}

@@

\begin{proposition}
\label{prop:rafStemp}
Let \((q, n) \in \left(\mathbb{N}^* \right)^2\) and let \(f\) be a coloring associated to an S-template with width \(q\) and \(n+1\) 
colors. Let \(b \in \mathbb{N}\) and assume there is a coloring \(g\) of 
\([\![1, b]\!]\) with \(n+1\) colors such that:

@@

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the coloring \(x \longmapsto g(x - q S(n))\), we have
\[ S(n+k) \geqslant q S(k) + b.\]
\end{proposition}
This proposition corresponds to the fact that the hypotheses made on the coloring of the last row can be weakened because 
the constraints in the definition of an S-template prevent a number from: 
\begin{itemize}
\item being the sum of two of numbers of its own color. 
\item creating a sum in its own color. 
\end{itemize}
The later is not always relevant on the last row given that it is not completely filled.
As a result, we can change the coloring on \([\![ q S(k) + 1,  q S(k) + b]\!]\) (the last row) to extend the previous partitions.
\newline

@@

\begin{theorem}
	Let \((p,k), (q,n) \in (\mathbb{N}^*)^2\). If there are an S-template with width \(q\) and \(n+1\) colors,
	and an S-template with width \(p\) and \(k\) colors, then there also is an S-template with width \(pq\) and \((n+k)\) 
	colors.
\end{theorem}

@@

An S-template with \(n\) colors is said to be symmetric if the sum-free \(n\)-partition derived 
from this template by applying the extension procedure of Theorem \ref{thm:Stemp} with a sum-free 
partition with length one is symmetric. 
\end{definition}

@@

\par
Inequalities (\ref{S(n+1)}), (\ref{S(n+2)}) and (\ref{S(n+3)}) cannot be further improved (with this definition of S-template). 
Inequality (\ref{S(n+4)}) cannot be further improved by only searching for symmetric S-templates whose special color is the 
last in the order of apparition (and with a multiplicative factor less than or equal to 118). Inequality (\ref{S(n+5)}) can most 
likely be further improved but the improvement probably will not be substantial. Finally, an S-template corresponding to inequality 
(\ref{S(n+6)}) was found by extending into an S-template the Schur 6 partition used in \cite{rowley2021improved} (owing to the 
size of this template, its interest is limited and since it can easily be derived from above mentioned partition, it is not given in 
\ref{S-templates}). Although we could not find a better S-template with seven colors, inequality (\ref{S(n+6)}) can definitely be 
improved by a wide margin.

### 2021-10-14 article/weakschur.tex
modifs :O

First, we put the integers of \([\![1, ap + q]\!]\) in the following table (with \(a\) columns and \(p + 1\) lines)
and divide it into 3 blocks (the columns are numbered from \(-l\) to \(+q\)).

@@

Note that with this numbering of columns, the column of the sum of two numbers is the only integer in \([\![-l,q]\!]\) equal to the
sum of the columns modulo \(a\).

@@

\noindent \underline{\textbf{\(\mathcal{C}\) block}}
\par
In this block, we use colors \(1, ..., n\). We color an integer \(x\) according to its projection on the first row \(\pi(x)\) offeset by \(a\), 
that is its column number. A simple solution is to color \(x\) with the same color as \(\pi(x) - a\) in the weakly sum-free coloring of 
\([\![1,q]\!]\). As long as \(2q \leqslant a + r\) (not going two far to the right) and there is no \(x \in \pi(\mathcal{C}) - a\) such that 
\(2x \in \pi(\mathcal{C}) - a\) (so that we do not have a sum in \(\mathcal{C}\) when taking two numbers in the same column), 
the colors \(1, ..., n\) are sum-free.

Setting \(w = l = \left\lceil \displaystyle \frac{q}{2} \right\rceil\) and \(r = \left\lfloor \displaystyle \frac{q}{2} \right\rfloor\) satsfies these 
constraints: this is the construction corresponding to Theorem \ref{thm:WS-ab}. \\

@@

\begin{definition}
Let \( n \in \mathbb{N}^*\). We define :
\[
\WS^+(n)=\max_{b\in \mathbb{N}^*} \WS_b^+(n).
\]
\end{definition}
The following proposition briefly shows how \(\WS^+(n)\) compares to weak Schur numbers.
\begin{proposition}
Let \(n \in [\![2, +\infty]\!]\). Then :
\[
\frac{3}{2} \WS (n-1)+1 \leqslant \WS^+(n) \leqslant \WS (n).
\]
\end{proposition}

@@

Theorem \ref{thm:Stemp} describes a manner to extend a S-template into a larger Schur partitions using other Schur 
partitions. Likewise, a WS-template can be extended into larger weak Schur partitions by using Schur partitions. This 
is the object of the following theorem.

@@

\par
NB: To keep the notation short, the conditions \(x + y \leqslant p\)  and \(x + y \leqslant a + b\) are omitted in the following five predicates.
\par
The (weakly) sum-free conditions are expressed as:
\begin{equation}
\forall (x,y) \in [\![1,a + b]\!]^2, \left\{
\begin{array}{l}
	f(x) = f(y) \\
	x \neq y
\end{array}
\right. \Longrightarrow f(x+y) \neq f(x),
\end{equation}
\begin{equation}
\forall (x,y) \in [\![b+1,a + b]\!]^2, f(x) = f(y) \Longrightarrow f(x+y) \neq f(x),
\end{equation}
\begin{equation}
\forall (x,y) \in [\![1,p]\!]^2, g(x) = g(y) \Longrightarrow g(x+y) \neq g(x).
\end{equation}

The additionnal constraints for the WS-template are:
\begin{equation}
\forall (x,y) \in [\![1,a + b]\!]^2, \left\{
\begin{array}{l}
	f(x) = f(y) \leqslant n \\
	x + y > a + b
\end{array}
\right. \Longrightarrow f(\pi(x+y)) \neq f(x),
\end{equation}
\begin{equation}
\forall (x,y) \in [\![1,a + b]\!]^2, \left\{
\begin{array}{l}
	f(x) = f(y) = n + 1 \\
	x + y > 2 a + b
\end{array}
\right. \Longrightarrow f(x+y - 2 a) \neq f(x).
\end{equation}

@@

\begin{proof}[\textsc{Proof of Theorem \ref{thm:WS-ab}.}]
\label{PreuveThm}
Let \((q, n) \in (\mathbb{N}^*)^2\) such that there is a partition of \([\![1, q]\!]\) into n weakly sum-free subsets. 
Let \(f: [\![1, q]\!] \rightarrow [\!]1, n]\!]\) a weakly sum-free colouring.
Let \(b=q\) and \(a = q + \left \lceil \dfrac{q}{2} \right \rceil + 1\). A new colouring \(g\) is defined as follows:

@@

\begin{theorem}
\label{thm:compWStemp}
Let \((k,p) \in (\mathbb{N}^*)^2\) and \((a, n, b) \in (\mathbb{N}^*)^3\). If there are a S-template with width
\(p\) and \(k+1\) colors and a \(b\)-WS-template with width \(a\) and \(n\) colors, then there is a \(pb\)-WS-template
with width \(pq\) and \((n+k)\) colors.
\end{theorem}

@@

Inequality (\ref{WS(n+3)}) cannot be further improved and was found with a SAT solver. It uses the first 
sophistication explained in Subsection \ref{ConstructionWS} in order to add
the last number in the first color. As for inequality (\ref{WS(n+4)}), it was obtained by combining a S-template with 
width 33 with a WS-template with width 4. The best template we could find with a computer search gives the 
inequality \(\WS(n+4) \geqslant 127 S(n) + 68\). It was also found with the SAT solver. In order to reduce the search space, 
we only looked for WS-templates with five colors which start with a near-optimal \(\WS(4)\) partition and we assumed that the 
special color was the last by order of appearance. We highly suspect that better WS-templates with \(n \geqslant 5\) 
colors can be found but one would have not to use the above assumptions.

>>>>

First, we construct the following table with \(a\) columns and \(p+1\) rows containing the integers from \([\![1, ap + q]\!]\). 
The columns are numbered from \(-l\) to \(+q\). This table is made of three different blocks.

@@

@@

\noindent \underline{\textbf{\(\mathcal{C}\) block}}
\par
In this block, we use colors \(1, ..., n\). We color an integer \(x\) according to its projection on the first row \(\pi(x)\) offeset by \(a\), 
that is its column number. A simple solution is to color \(x\) with the same color as \(\pi(x) - a\) in the weakly sum-free coloring of 
\([\![1,q]\!]\). As long as \(2q \leqslant a + r\) (not going two far to the right) and there is no \(x \in \pi(\mathcal{C}) - a\) such that 
\(2x \in \pi(\mathcal{C}) - a\), we do not have a sum in \(\mathcal{C}\) when taking two numbers in the same colum. Finally, the 
weakly sum-free property of the coloring of \([\![1,q]\!]\) guarantees that if \(x \in [\![1,r]\!]\) and \(y \in \mathcal{C}\) are colored 
the same, then \(x + y\) is colored differently.

Setting \(w = l = \left\lceil \displaystyle \frac{q}{2} \right\rceil\) and \(r = \left\lfloor \displaystyle \frac{q}{2} \right\rfloor\) satsfies these 
constraints: it corresponds to Theorem \ref{thm:WS-ab}. \\

@@

\begin{definition}
Let \( n \in \mathbb{N}^*\). We define :
\[
\WS^+(n)=\max_{b\in \mathbb{N}^*} \WS_b^+(n).
\]
\end{definition}
The following proposition shows how \(\WS^+(n)\) compares to weak Schur numbers.
\begin{proposition}
Let \(n \in [\![2, +\infty]\!]\). Then :
\[
\frac{3}{2} \WS (n-1)+1 \leqslant \WS^+(n) \leqslant \WS (n).
\]
\end{proposition}

@@

Theorem \ref{thm:Stemp} describes a manner to extend an S-template into a larger Schur partitions using other Schur 
partitions. Likewise, a WS-template can be extended into larger weak Schur partitions by using Schur partitions. This 
is the object of the following theorem.

@@

\par
NB: To keep the notation short, the conditions \(x + y \leqslant p\)  and \(x + y \leqslant a + b\) are omitted in the following five predicates.

The sum-free conditions are expressed as:
\[
\forall (x,y) \in [\![b+1,a + b]\!]^2, f(x) = f(y) \Longrightarrow f(x+y) \neq f(x),
\]
\[
\forall (x,y) \in [\![1,p]\!]^2, g(x) = g(y) \Longrightarrow g(x+y) \neq g(x).
\]

The weakly sum-free condition is expressed as:
\[
\forall (x,y) \in [\![1,a + b]\!]^2, \left\{
\begin{array}{l}
	f(x) = f(y) \\
	x \neq y
\end{array}
\right. \Longrightarrow f(x+y) \neq f(x),
\]

The additionnal constraints for the WS-template are expressed as:
\[
\forall (x,y) \in [\![1,a + b]\!]^2, \left\{
\begin{array}{l}
	f(x) = f(y) \leqslant n \\
	x + y > a + b
\end{array}
\right. \Longrightarrow f(\pi(x+y)) \neq f(x),
\]
\[
\forall (x,y) \in [\![1,a + b]\!]^2, \left\{
\begin{array}{l}
	f(x) = f(y) = n + 1 \\
	x + y > 2 a + b
\end{array}
\right. \Longrightarrow f(x+y - 2 a) \neq f(x).
\]

@@

\begin{proof}[\textsc{Proof of Theorem \ref{thm:WS-ab}.}]
\label{PreuveThm}
Let \((q, n) \in (\mathbb{N}^*)^2\) such that there is a partition of \([\![1, q]\!]\) into n weakly sum-free subsets. 
Let \(f: [\![1, q]\!] \rightarrow [\!]1, n]\!]\) be a weakly sum-free colouring.
Let \(b=q\) and \(a = q + \left \lceil \dfrac{q}{2} \right \rceil + 1\). A new colouring \(g\) is defined as follows:

@@

\begin{theorem}
\label{thm:compWStemp}
Let \((k,p) \in (\mathbb{N}^*)^2\) and \((a, n, b) \in (\mathbb{N}^*)^3\). If there are an S-template with width
\(p\) and \(k+1\) colors and a \(b\)-WS-template with width \(a\) and \(n\) colors, then there is a \(pb\)-WS-template
with width \(pq\) and \((n+k)\) colors.
\end{theorem}

@@

Inequality (\ref{WS(n+3)}) cannot be further improved and was found with a SAT solver. It uses the first 
sophistication explained in Subsection \ref{ConstructionWS} in order to add
the last number in the first color. As for inequality (\ref{WS(n+4)}), it was obtained by combining an S-template with 
width 33 with a WS-template with width 4. The best template we could find with a computer search gives the 
inequality \(\WS(n+4) \geqslant 127 S(n) + 68\). It was also found with the SAT solver. In order to reduce the search space, 
we only looked for WS-templates with five colors which start with a near-optimal \(\WS(4)\) partition and we assumed that the 
special color was the last by order of appearance. We highly suspect that better WS-templates with \(n \geqslant 5\) 
colors can be found but one would have not to use the above assumptions.
